---
title: "Reproducible research: Peer Assignment: 1"
output: html_document
---

The code requires the file **activity.zip** in the working directory.

<h2> 1. Loading and preprocessing the data </h2>

```{r, echo=TRUE}
library(dplyr)
if (!file.exists("activity.csv")) {unzip("activity.zip")} # unzip only if not done yet

activity<-read.csv("activity.csv", header=TRUE, na="NA") 
activity$date<-as.Date(activity$date) # change format of date values
activity$steps<-as.numeric(activity$steps) # change format of date values
activity$interval<-as.numeric(activity$interval) # change format of date values
head(activity)
```

<h2> 2. What is mean total number of steps taken per day? </h2>

Missing values are ignored in this step.
The histogram of the total number of steps taken each day:
```{r, echo=TRUE}
activity_day<-aggregate(steps ~ date, activity, sum)
barplot(activity_day$steps, names.arg=activity_day$date,las=1,main="Total steps per day", srt = 60,xlab="date", ylab="steps", col="blue")
```

Calculated mean and median total number of steps taken per day:
```{r, echo=TRUE}
mean(activity_day$steps, na.rm=TRUE) # mean
median(activity_day$steps, na.rm=TRUE) # median
```

Time series plot of 5-minute interval (x-axis) and the average number of steps taken.
<h2> What is the average daily activity pattern? </h2>
```{r, echo=TRUE}
activity_ts<-activity %.% group_by(interval) %.% summarise(meanInterval = mean(steps, na.rm=TRUE))
plot(activity_ts, type="l", main="Mean number of steps by interval", srt = 60,xlab="interval", ylab="mean steps", col="blue")
```

Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r, echo=TRUE}
activity_ts[which.max(activity_ts$meanInterval),1]
           
```

<h2>Imputing missing values</h2>
Calculate and report the total number of missing values in the dataset. Then, create a new data frame based on the original activity data by filling the missing values with the mean as calculated above for the interval plot.
```{r, echo=TRUE}
sum(!complete.cases(activity$steps))
activity_filled <- activity
n<-nrow(activity)
for (r in 1:n) {
    if (is.na(activity_filled$steps[r]))
        {
            activity_filled$steps[r]<-activity_ts$meanInterval[activity_ts$interval==activity$interval[r]]
        }
}

```

Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day.

```{r, echo=TRUE}
activity_day_filled<-aggregate(steps ~ date, activity_filled, sum)
barplot(activity_day_filled$steps, names.arg=activity_day_filled$date,las=1,main="Total steps per day", srt = 60,xlab="date", ylab="steps", col="blue")
```

Calculated mean and median total number of steps taken per day:
```{r, echo=TRUE}
mean(activity_day_filled$steps) # mean
median(activity_day_filled$steps) # median
```
The impcat using the above adjustment for NA's is low.

```{r, echo=TRUE}
step_diff <- sum(activity_day$steps) - sum(activity_day_filled$steps)
```

<h2>Are there differences in activity patterns between weekdays and weekends?</h2>

The plot shows that on weekends, activity in the morning starting around 5:30 is higher than on weekdays. Both time series have a maximum in the morning around 9:00. During the day, 10:00 to 18:00, activity is lower on weekends. At night, activity is close to zero for both.
```{r, echo=TRUE}
week <- factor(weekdays(activity_filled$date) %in% c("Saturday","Sunday"), 
               labels=c("weekday","weekend"))

activity_dayType <- aggregate(activity_filled$steps, by=list(interval=activity_filled$interval, 
                                                             weekday=week), mean)

activity_weekend <- subset(activity_dayType,weekday=="weekday")
activity_weekday <- subset(activity_dayType,weekday=="weekend")


plot(activity_weekend$interval/max(activity_weekend$interval)*24,activity_weekend$x, type="l", col="blue", xlab="time", ylab="avg number of steps in each 5 min interval", lwd=3,xaxt = "n")
axis(1, at=0:24, labels=paste(0:24,":00", sep=""))

lines(activity_weekday$interval/max(activity_weekday$interval)*24, activity_weekday$x, col="red", lwd=3)

legend("topright",lty = c(1,1), 
        col = c("blue", "red"), 
        legend = c("weekend", "weekday"))

```


```{r, echo=TRUE, fig.height=10}
par(mfrow=c(2,1))
plot(activity_weekend$interval/max(activity_weekend$interval)*24,activity_weekend$x, type="l", col="blue", xlab="time", ylab="steps", lwd=3,xaxt = "n", main="weekend")
axis(1, at=0:24, labels=paste(0:24,":00", sep=""))

plot(activity_weekday$interval/max(activity_weekday$interval)*24,activity_weekday$x, type="l", col="red", xlab="time", ylab="steps", lwd=3,xaxt = "n", main="weekday")
axis(1, at=0:24, labels=paste(0:24,":00", sep=""))

```

